The role of cognateness in native spoken word recognition

### 2024-10-15 manuscript/manuscript.qmd
Add General discussion

The present work investigated the dynamics of lexical activation and selection dring non-native speech listening. Using a translation elicitation task, we collected data from a group of English native adults (Experiment 1) and a group of Spanish native adults (Experiment 2). Neither group was familiar with Catalan, Spanish, or any other Romance language. In each trial, participants were presented with Spanish words (Experiment 1) or Catalan words (Experiments 1 and 2), and then asked to type their best-guess translation for the presented word in their native language. Participants provided correct translations more often if the presented word and its translation shared phonological similarity, only if the number of phonological neighbours around the presented word was small. These results suggest that listening to words in an unfamiliar language can trigger dynamics of lexical activation and selection similar to those triggered by listening to speech in the native language.

In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses, despite reporting no prior relevant experience with the Catalan or Spanish. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are two Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close or distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant's performance. Some participants in Experiments 1 and 2 were surprisingly good at translating a subset of words which had low phonological similarity with their correct translation. We were concerned that this may be caused by some prior knowledge of specific words by our participants, as some Spanish words are commonly seen in media or product labels, making them familiar even to monolingual speakers of English. In Experiment 3, we collected additional data from a new group of British English natives. The design was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. After removing responses in which participants reported prior knowledge about the meaning of the presented Catalan or Spanish word, we found similar resutls as in Experiment 1, suggesting that participants did not simply rely on prior familiarity with the word-forms presented but rather exclusively on their phonological similarity with the correct English translation.

Our findings provide insights into the mechanisms underlying homophonic translations. This phenomenon is frequently experienced by listeners of foreign speech, when strings of sounds in the input are perceived as strings of words or phrases in the native language, without necessarily preserving the meaning.

This type of cross-linguistic similarity has also been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory].

>>>>

The present work explored the psycholinguistic bases of homophonic translation, a phenomenon in which listening to speech in a non-native---perhaps unfamiliar---language leads to the activation of lexical representations in the native language, without necessarily preserving the meaning. We investigated how phonological similarity and its interaction with phonological neighbourhood density impact the dynamics of lexical activation and selection during non-native word processing. We designed a translation elicitation task in which participants listened to individual words in a non-native, unfamiliar language. After listening to each word, participants were asked to provide their best-guess translation for each word in their native language. We run three experiments. In Experiment 1, British English-native adults listened to a list of words in Catalan or Spanish. Participants reported no prior familiarity with Catalan, Spanish, or any other Romance language; yet, participants provided accurate translations for words that shared some degree of phonological similarity with their correct English translation. Participants only benefited from phonological similarity when the number of English phonological neighbours of the presented word was small. Experiment 2 was aimed at extending our findings to a population of participants whose native language was typologically closer to the unfamiliar presented language. We tested a group of Spanish native adults who listened to a list of Catalan words. Again, participants reported no prior familiarity with Catalan, or any other Romance language other than Spanish. In line with the results in English natives, we also found an interaction between phonological similarity and phonological neighbourhood size. The facilitation effect of phonological similarity in the Spanish participants was more robust to the interferring effect of phonological neighbourhood density, providing correct translations for words with high neighbourhood densities more often than English participants. In Experiment 3, we replicated the results from Experiment 1, while additionally controlling for participants' prior familiarity with the Catalan and Spanish words. We suspected that some English participants might have managed to surprisingly provide correct translations for some low-phonological similarity words---despite having reported being unfamiliar with Catalan, Spanish, or any other Romance language---thanks to prior experience with those words. We collected additional data from a new group of British English native adults. The experimental design in Experiment 3 was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. After removing responses in which participants reported prior knowledge about the meaning of the presented Catalan or Spanish word, we found similar results as in Experiment 1, suggesting that participants did not simply rely on prior familiarity with the word-forms presented but rather exclusively on their phonological similarity with the correct English translation.

Our findings provide novel psycholinguistic insights into homophonic translation from an experimental perspective. Homophonic is a frequent, often spontaneous phenomenon that has a has been described across many cultures and  languages. The most famous term coined for it is *Soramimi*, in Japanese. *Soramimi* refers specifically to homophonic translation in a musical context, when particular sections of the lyrics of a song in a non-native language make listeners think of particular words and phrases in their own native language, with possibly a completely different meaning. While *Somoramimi* refers to the spontaneous ocurrence of homophonic translations, this phenomenon is often intentionally exploited by translators to preserve the meter and "sound structure" (e.g., rhymes) of the original poem in the target language [e.g.,@levick2016translating;  @pilshchikov2016semiotics], in such way that the translator "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For isntance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).

The underlying mechanisms that lead to spontaneous homophonic translations are informative of the undeerlying mechanisms of native speech processing, yet remain relatively unexplored. How does the lexical system deal with the non-native speech signal, in which form-meaning associations are scarce, if not absent? The present work addressed this issue by exploring the conditions under which non-native speech is likely to result in correct translations. To account for the fact that prior familiarity with the language (e.g., form-meaning associations) may contribute to correct translations, we tested participants in a language that they reported being unfamiliar with. In such conditions, participants can almost exlusively rely on the acoustic properties of the speech signal to provide correct translations.

Our findings suggest that participants were able to exploit the phonological similarity between the presented words and their correct translation, even when both words shared less than 50% of their phonological form (as measured by their Levenshtein distance). How the perceptual system deals with non-native phonemes is a vastly studied subject. The same cannot be said about how the perceptual system deals with non-native *word-forms*. The present study provides experimental evidence suggesting that participants accommodate non-native phonological forms to their native phoneme inventory, leading to the activation of native lexical representations. One of the limitations of the present study is that Levenshtein similarity underestimates the actual phonetic similarity between the signal and the phonological form of the correct translations. Phonological transcriptions like X-SAMPA---which we used to calculate Levenshtein similarity----are abstract representations that ignore non-phonological contrasts (e.g., phones), and consider different symbols as completely different phonemes disregarding the fact that some phonemes share more phonetic features than others. Altogether, it is likely that participants exploited a more fine grained, raw source of information to provide correct answers.

Our findings also suggest that, in line with previous studies of word-processing, phonological neighbourhood size interfered participants' ability to produce correct translations, even for words with high phonological similarity with their correct translation. We calculated a measure of cross-linguistic neighbourhood size by counting, for each presented word (in Catalan or Spanish in Experiments 1 and 3, Catalan in Experiment 2) the number of phonologically related words (with a Levenshtein distance of one) in the target language (English in Experiments 1 and 3, Spanish in Experiment 2). To account for the fact that competition between phonological neighbours is sensitive to lexical frequency (high-freuqency neighbours produce stronger interference effects), we only counted neighbours with a lexical freuqency higher than the correct translation. Using this measure, we found that, across the three experiments, participants ability to exploit phonological similarity to produce correct translations declined as the number of phonological neighbours increased.

A comparison between results in Experiments 1 and 2 suggested that the performance of Spanish participants translating Catalan was more resilient to the interfering effects of phonological neighbourhood density than the performance of English participants translating Catalan or Spanish. As highlihted before, Catalan and Spanish (both Romance languages) are typologically closer to each other than to English (a Germanic language). Catalan and Spanish share a higher proportion of cognates (i.e., form-similar translation equivalents) than Catalan and English, or Spanish and English. Previous studies in bilinguals show that this kind of cross-linguistic similarity impacts lexical processing:  words that share high phonological similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory]. To our knowledge, this present study is the first to investigate the role of cross-linguistic phonological similarity in monolinguals. In particular, our participants were reportedly unfamiliar with the presented language. As such, they had to complete our translation elicitation task in the absence of form-meaning mappings, therefore relying exclusively on the phonological similarity between the presented words and their correct translations. This provides a convenient case to investigate the role of phonological similarity in lexical processing, disantangling the contribution of phonological similarity from the contribution of conceptual equivalence. Future studies might expand the methods used in the present study to elaborate in this issue.

Two populations of speakers were tested in the present study: British English native adults living in the United Kindgom, and Spanish native adults living in Spain. Our findings suggest that the typological distance between the presented and target language is associated with a more robust facilitation effect of phonological similarity. Whether such association is linear (i.e., robustness of facilitation increases linearly with typological distance), proportional, or even extends beyond the specific pair of languages involved in the present work remains unclear. The generalisability of these findings can also be tested by increasing the repertoire of words involved in th translation elicitation task. In Experiments 1 to 3, participants answered to a maximum of 105 words. All words were of high-frequency, and had a low age-of-acquisition. In order to characterise listeners' ability to translate correctly words from an unfamiliar language, words with varying levels of difficulty should be included.

In summary, the present paper provides insights into the psycholinguistic mechanisms underlying homophonic translation. English and Spanish native adults were tested in a translation elicitation task in which they had to guess the English or Spanish translation of a series of words in an unfamiliar language. Participants successfully exploited phonological similarity between the presented words and their correct translations to provide correct answers. Participants' performance in the task only benefited from phonological similarity when the presented word had few higher-frequency phonological neighbours in the target language. Finally, the faciltiation effect of phonological similarity was more robust in the Spanish native participants, who translated words from a typologically closer language than English participants. Overall, the findings presented in the present paper suggest that the processing of words in a non-native, unfamiliar language recruits mechanisms of lexical activation, selection, and interference parallel to those recruited by listening to words in a native language.

### 2024-10-15 manuscript/manuscript.tex
Add General discussion

From the 3,145 total responses provided by English participants who
listened to Catalan words, participants reported having prior knowledge
of the presented Catalan words in 446 (14.18\%) of them. In those
responses, participants reported an average of 5.05 confidence in the
0-8 scale (\emph{SD} = 1.94). In responses in which no prior knowledge
of the presented word was reported, average confidence was 1.13
confidence in the 0-8 scale (\emph{SD} = 1.38). From the 2,743 total
responses provided by English participants who listened to Spanish
words, participants reported having prior knowledge of the presented
Spanish words in 446 (14.18\%) of them. In those responses, participants
reported an average of 4.79 confidence in the 0-8 scale (\emph{SD} =
1.82). In responses in which no prior knowledge of the presented word
was reported, average confidence was 1.25 confidence in the 0-8 scale
(\emph{SD} = 1.66). Before data analysis, responses in which
participants reported prior knowledge about the meaning of the presented
Catalan or Spanish word were excluded from the dataset.

@@

The present work investigated the dynamics of lexical activation and
selection dring non-native speech listening. Using a translation
elicitation task, we collected data from a group of English native
adults (Experiment 1) and a group of Spanish native adults (Experiment
2). Neither group was familiar with Catalan, Spanish, or any other
Romance language. In each trial, participants were presented with
Spanish words (Experiment 1) or Catalan words (Experiments 1 and 2), and
then asked to type their best-guess translation for the presented word
in their native language. Participants provided correct translations
more often if the presented word and its translation shared phonological
similarity, only if the number of phonological neighbours around the
presented word was small. These results suggest that listening to words
in an unfamiliar language can trigger dynamics of lexical activation and
selection similar to those triggered by listening to speech in the
native language.

In Experiment 1, we collected data from two groups of British English
natives living in the United Kingdom. One group was presented with
Catalan words, the other with Spanish words. We examined the extent to
which participants were able to use the phonological similarity between
the presented word (in Catalan or Spanish) and its target translation to
provide accurate responses, despite reporting no prior relevant
experience with the Catalan or Spanish. In Experiment 2, we tested a
group of (European) Spanish natives in the same task, who were presented
with Catalan words. Catalan and Spanish are two Romance languages whose
close typological distance is reflected in the fact that they share many
cognates, where English is a Germanic language that shares considerably
fewer cognates with Catalan and Spanish. By testing participants
translating words from typologically close or distant languages, we
expected to widen the range of the phonological similarity scores of the
translation pairs involved in the experimental task, therefore allowing
us to explore potential cross-language differences in participant's
performance. Some participants in Experiments 1 and 2 were surprisingly
good at translating a subset of words which had low phonological
similarity with their correct translation. We were concerned that this
may be caused by some prior knowledge of specific words by our
participants, as some Spanish words are commonly seen in media or
product labels, making them familiar even to monolingual speakers of
English. In Experiment 3, we collected additional data from a new group
of British English natives. The design was closely modelled after
Experiment 1, except that after providing their response in each trial,
participants reported whether they had previous knowledge of the
presented word. After removing responses in which participants reported
prior knowledge about the meaning of the presented Catalan or Spanish
word, we found similar resutls as in Experiment 1, suggesting that
participants did not simply rely on prior familiarity with the
word-forms presented but rather exclusively on their phonological
similarity with the correct English translation.

Our findings provide insights into the mechanisms underlying homophonic
translations. This phenomenon is frequently experienced by listeners of
foreign speech, when strings of sounds in the input are perceived as
strings of words or phrases in the native language, without necessarily
preserving the meaning.

This type of cross-linguistic similarity has also been extremely
informative for the study of bilingual lexical processing: words that
share high form-similarity with their translations are processed faster
and more accurately than words sharing lower phonological similarity.
This phenomenon has provided strong evidence in favour of the
language-non selective hypothesis of bilingual lexical access, which
states that bilinguals activate lexical representations in both
languages, even during monolingual situations. This facilitation effect
has been reported for comprehension
(\citeproc{ref-dijkstra2010cross}{Dijkstra et al., 2010};
\citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}).

@@

@@

@@

>>>>

From the 3,145 total responses provided by English participants who
listened to Catalan words, participants reported having prior knowledge
of the presented Catalan words in 446 (14.18\%) of them. In those
responses, participants reported an average of 5.05 confidence in the
0-8 scale (\emph{SD} = 1.94). In responses in which no prior knowledge
of the presented word was reported, average confidence was 1.13
confidence in the 0-8 scale (\emph{SD} = 1.38). From the 2,743 total
responses provided by English participants who listened to Spanish
words, participants reported having prior knowledge of the presented
Spanish words in 197 (7.18\%) of them. In those responses, participants
reported an average of 4.79 confidence in the 0-8 scale (\emph{SD} =
1.82). In responses in which no prior knowledge of the presented word
was reported, average confidence was 1.25 confidence in the 0-8 scale
(\emph{SD} = 1.66). Before data analysis, responses in which
participants reported prior knowledge about the meaning of the presented
Catalan or Spanish word were excluded from the dataset.

@@

The present work explored the psycholinguistic bases of homophone
translations, a phenomenon in which listening to speech in a
non-native---perhaps unfamiliar---language leads to the activation of
lexical representations in the native language. We investigated how
phonological similarity and its interaction with phonological
neighbourhood density impact the dynamics of lexical activation and
selection during non-native word processing. We designed a translation
elicitation task in which participants listened to individual words in a
non-native, unfamiliar language, and then were asked to provide their
best-guess translation for each word in their native language. We run
three experiments. In Experiment 1, British English-native adults
listened to a list of words in Catalan or Spanish. Participants reported
no prior familiarity with Catalan, Spanish, or any other Romance
language; yet, participants provided accurate translations for words
that shared some degree of phonological similarity with their correct
English translation. Participants only benefited from phonological
similarity when the number of English phonological neighbours of the
presented word was small. Experiment 2 was aimed at extending our
findings to a population of participants whose native language was
typologically closer, to the (still unfamiliar) presented language. We
tested a group of Spanish native adults who listened to a list of
Catalan words. Again, participants reported no prior familiarity with
Catalan, or any other Romance language other than Spanish. In line with
the results in English natives, Spanish natives also showed an
interaction between phonological similarity and phonological
neighbourhood size. However, the facilitation effect of phonological
similarity in the Spanish participants was more robust to the
interferring effect of phonological neighbourhood density, providing
correct translations for high-neighbourhood density words more often
than participants in Experiment 1. In Experiment 3, we replicated the
results from Experiment 1, while controlling for participants' prior
familiarity with the Catalan and Spanish words. We suspected that some
English participants might have managed to provide correct translations
for some low-phonological similarity words despite being reportedly
unfamiliar with Catalan, Spanish, or any other Romance language, thanks
to previous exposure to those words. We collected additional data from a
new group of British English natives. The design was closely modelled
after Experiment 1, except that after providing their response in each
trial, participants reported whether they had previous knowledge of the
presented word. After removing responses in which participants reported
prior knowledge about the meaning of the presented Catalan or Spanish
word, we found similar resutls as in Experiment 1, suggesting that
participants did not simply rely on prior familiarity with the
word-forms presented but rather exclusively on their phonological
similarity with the correct English translation.

The results in the present work tap into homophonic translation, a
frequent phenomenon to many speakers to occurs when speech in a
non-native language activates lexical representations in the native
language withhout necessarily preserving the meaning. This phenomenon
has been described across many languages, and coined with different
terms, the most famous being the \emph{Soramimi} in Japanese.
\emph{Soramimi} refers specifically to homophonic translation in the
context of music, when particular sections of the lyrics of a song in a
non-native language make listeners think of particular words and phrases
in their own native language, with possibly a completely different
meaning. While \emph{Somoramimi} refers to the spontaneous ocurrence of
homophonic translations, homophonic translation is often exploited by
translators to preserve the meter and ``sound structure'' (e.g., rhymes)
of the original poem in the target language (e.g.,
\citeproc{ref-levick2016translating}{Levick, 2016};
\citeproc{ref-pilshchikov2016semiotics}{Pilshchikov, 2016}), in such way
that the translator ``sacrifice{[}s{]} fidelity to the image rather than
the melodiousness of verse'' (\citeproc{ref-briusov1911ot}{Briusov,
1911}). In the same line, Edith Grossman's translation of \emph{El
Quijote} (\citeproc{ref-cervantes}{\textbf{cervantes?}}) present a
similar case, in which formal aspects of the original text in Spanish
are favoured to some degree over equivalence to English, even sometimes
keeping Spanish words from the original text in the English translation
(e.g., \emph{Señor}, \emph{ínsula}).

The underlying mechanisms that lead to spontaneous homophonic
translations are informative of the undeerlying mechanisms of native
speech processing: how does the lexical system deal with the non-native
speech signal, in which form-meaning associations are scarce, if not
absent? This question remains relatively explored from a
psycholinguistic standpoint. The present work was addressed at exploring
the conditions under which non-native speech is likely to result in
correct translations. To account for the fact that prior familiarity
with the language (e.g., form-meaning associations) may contribute to
correct translations, we tested participants in a language that they
reported being unfamiliar with. In such conditions, participants can
almost exlusively rely on the acoustic properties of the speech signal
to provide correct translations.

Our findings suggest that participants were able to exploit the
phonological similarity between the presented words and their correct
translation, even when both words shared less than 50\% of their
phonological form (as measured by their Levenshtein distance). How the
perceptual system deals with non-native phonemes is a vastly studied
subject. The same cannot be said about how the perceptual system deals
with non-native word-forms. We provide new experimental evidence
suggesting that participants accommodate non-native phonological forms
to their native phoneme inventory, leading to the activation of native
lexical representations. One of the limitations of the present study is
that Levenshtein similarity underestimates the actual phonetic
similarity between the signal and the phonological form of the correct
translations. Phonological transcriptions like X-SAMPA---which we used
to calculate Levenshtein similarity----are abstract representations that
ignore non-phonological contrasts (e.g., phones), and consider different
symbols as completely different phonemes disregarding the fact that some
phonemes share more phonetic features than others. Altogether, it is
likely that participants exploited a more fine grained, raw source of
information to provide correct answers.

Our findings also suggest that, in line with previous studies of
word-processing, phonological neighbourhood size interfered
participants' ability to produce correct translations, even for words
with high phonological similarity with their correct translation. We
calculated a measure of cross-linguistic neighbourhood size by counting,
for each presented word (in Catalan or Spanish in Experiments 1 and 3,
Catalan in Experiment 2) the number of phonologically related words
(with a Levenshtein distance of one) in the target language (English in
Experiments 1 and 3, Spanish in Experiment 2). To account for the fact
that competition between phonological neighbours is sensitive to lexical
frequency (high-freuqency neighbours produce stronger interference
effects), we only counted neighbours with a lexical freuqency higher
than the correct translation. Using this measure, we found that, across
the three experiments, participants ability to exploit phonological
similarity to produce correct translations declined as the number of
phonological neighbours increased.

A comparison between results in Experiments 1 and 2 suggested that the
performance of Spanish participants translating Catalan was more
resilient to the interfering effects of phonological neighbourhood
density than the performance of English participants translating Catalan
or Spanish. As highlihted before, Catalan and Spanish (both Romance
languages) are typologically closer to each other than to English (a
Germanic language). Catalan and Spanish share a higher proportion of
cognates (i.e., form-similar translation equivalents) than Catalan and
English, or Spanish and English. Previous studies in bilinguals show
that this kind of cross-linguistic similarity impacts lexical
processing: words that share high phonological similarity with their
translations are processed faster and more accurately than words sharing
lower phonological similarity. This phenomenon has provided strong
evidence in favour of the language-non selective hypothesis of bilingual
lexical access, which states that bilinguals activate lexical
representations in both languages, even during monolingual situations.
This facilitation effect has been reported for comprehension
(\citeproc{ref-dijkstra2010cross}{Dijkstra et al., 2010};
\citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}). To
our knowledge, this present study is the first to investigate the role
of cross-linguistic phonological similarity in monolinguals. In
particular, our participants were reportedly unfamiliar with the
presented language. As such, they had to complete our translation
elicitation task in the absence of form-meaning mappings, therefore
relying exclusively on the phonological similarity between the presented
words and their correct translations. This provides a convenient case to
investigate the role of phonological similarity in lexical processing,
disantangling the contribution of phonological similarity from the
contribution of conceptual equivalence. Future studies might expand the
methods used in the present study to elaborate in this issue.

Two populations of speakers were tested in the present study: British
English native adults living in the United Kindgom, and Spanish native
adults living in Spain. Our findings suggest that the typological
distance between the presented and target language is associated with a
more robust facilitation effect of phonological similarity. Whether such
association is linear (i.e., robustness of facilitation increases
linearly with typological distance), proportional, or even extends
beyond the specific pair of languages involved in the present work
remains unclear. The generalisability of these findings can also be
tested by increasing the repertoire of words involved in th translation
elicitation task. In Experiments 1 to 3, participants answered to a
maximum of 105 words. All words were of high-frequency, and had a low
age-of-acquisition. In order to characterise listeners' ability to
translate correctly words from an unfamiliar language, words with
varying levels of difficulty should be included.

In summary, the present paper provides insights into the
psycholinguistic mechanisms underlying homophonic translation. English
and Spanish native adults were tested in a translation elicitation task
in which they had to guess the English or Spanish translation of a
series of words in an unfamiliar language. Participants successfully
exploited phonological similarity between the presented words and their
correct translations to provide correct answers. Participants'
performance in the task only benefited from phonological similarity when
the presented word had few higher-frequency phonological neighbours in
the target language. Finally, the faciltiation effect of phonological
similarity was more robust in the Spanish native participants, who
translated words from a typologically closer language than English
participants. Overall, the findings presented in the present paper
suggest that the processing of words in a non-native, unfamiliar
language recruits mechanisms of lexical activation, selection, and
interference parallel to those recruited by listening to words in a
native language.

@@

\bibitem[\citeproctext]{ref-briusov1911ot}
Briusov, V. Y. (1911). \emph{Ot perevodchika}. Skorpion.

@@

\bibitem[\citeproctext]{ref-levick2016translating}
Levick, T. (2016). Translating homophonic wordplay in patrick goujon's
moi non: A case study. \emph{Sound/Writing: On Homophonic Translation}.

@@

\bibitem[\citeproctext]{ref-pilshchikov2016semiotics}
Pilshchikov, I. (2016). The semiotics of phonetic translation.
\emph{Studia Metrica Et Poetica}, \emph{3}(1), 53--104.
